Clinical prediction using similarity cohort--based models: A systematic scoping review

### 2025-03-07 docs/Localized-Modeling-BMCAI/localized-modeling-dai-revision.tex
fix typos & update latex source

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

\title{Clinical prediction with localized modeling using
similarity-based cohorts: A scoping review}

\author{
    Adi Cohen
   \\
    College of Osteopathic Medicine, Nova Southeastern University \\
  Davie, FL \\
  \texttt{\href{mailto:ac4569@mynsu.nova.edu}{\nolinkurl{ac4569@mynsu.nova.edu}}} \\
   \And
    Patti McCall-Junkin
   \\
    Smathers Libraries Academic Research \& Consulting Services,
University of Florida \\
  Gainesville, FL \\
  \texttt{\href{mailto:pattimccall.junkin@gmail.com}{\nolinkurl{pattimccall.junkin@gmail.com}}} \\
   \And
    Jason Cory Brunson
   \\
    Laboratory for Systems Medicine, University of Florida \\
  Gainesville, FL \\
  \texttt{\href{mailto:jason.brunson@medicine.ufl.edu}{\nolinkurl{jason.brunson@medicine.ufl.edu}}} \\
  }

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{lscape}
\newcommand{\landscapebegin}{\begin{landscape}}
\newcommand{\landscapeend}{\end{landscape}}
\begin{document}
\maketitle

\begin{abstract}
Background: Applications of classical case-based reasoning (CBR) have
given rise to a family of techniques we call ``localized models'', in
which a statistical model is fitted to a neighborhood of labeled cases
matched by similarity to a target case. We aim to describe clinical and
health applications of localized models to date and propose a general
framework for their design and evaluation.

Methods: We searched four bibliographic platforms during 2021 July
19--22, updated 2024 January 24. We set four eligibility criteria to
identify applications of localized models to clinical and health tasks.
Two authors divided title/abstract screening and reviewed screened
entries for inclusion. We discussed settings, tasks, and tools;
identified and tabulated themes; and synthesized the methods into a
general framework.

Results: Of 1,657 search results, 360 were reviewed, then combined with
43 publications that seeded the review and 1 obtained by citation
tracking. 27 were included, published 1997--2022. The specificity of
search terms was poor, and inter-rater reliability was low. Almost all
models were predictive, the most common tasks being prognosis and
diagnosis. Most studies used clinical, occasionally laboratory and
image, data. Several addressed memory and runtime costs. A general
technique that specializes to most of those reviewed involved matching,
retrieval, fitting, and evaluation steps that could optionally be
supervised, optimized, or recursively performed.

Conclusions: Localized models have potential to improve the performance
of clinical decision support tools while maintaining interpretability,
but rigorous comparisons to competing methods must be conducted and
computational hurdles must be overcome. We hope that our review will
spur future work on efficiency, reproducibility, and user needs.
\end{abstract}

\keywords{
    case-based reasoning
   \and
    localized modeling
   \and
    nearest neighbors
   \and
    clinical decision support
   \and
    scoping review
  }

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

A main driver of the development of clinical information systems (CIS)
and clinical decision support (CDS) tools has been to leverage the
advantages of routinely collected health data toward improving care.
These advantages include their immediate accessibility through the
institutional EHR, billing database, or other source; their specificity
to the institution and the population it serves; and the closeness in
time of the available data. Though by definition not collected for
research use, the secondary research use of routinely collected health
data has been put forth, as ``practice-based evidence'', a complement to
the paradigm of evidence-based practice.

Over the same period of time, interest in individualizing care from
population-derived evidence-based recommendations to specific patient
needs has driven the application of advanced computational tools,
including artificial intelligence (AI). Whereas classical models
produced formulae involving a limited set of data elements that would be
uniformly applied to all patients, AI models often process hundreds or
more variables in opaque ways to yield predictions that depend on not
only their values but their associations and interplay with each other.
The loss of straightforward interpretations of these models has limited
their practical uptake and motivated the construction of explanatory
statistics for opaque models as well as the development of more
interpretable complex models.

One of few methodologies to address all three of these established needs
is one we call \emph{localized modeling}, which appears to have been
introduced several times independently under different names, and in
slightly different forms. The approach is a specialized form of
case-based reasoning (CBR) that relies on a patient similarity
measure\footnote{Similarity measures between more granular units of
  analysis, such as encounters and decision points, are often still
  referred to as patient similarity measures.} to extract a cohort of
past or training cases that then inform the diagnosis, prognosis, or
care of a new or test case. Traditional CBR returns these retrieved
cases to inform human decision-making; the related nearest neighbors
(NN) technique generates predictions from cohorts automatically via
averaging (regression) or voting (classification) of their outcomes. In
contrast, localized modeling fits families of predictive models, for
example generalized linear models (GLMs), to retrieved similarity
cohorts in order to generate predictions. Localized models thus provide
an individualized way to fit fully interpretable models to large
population data.

This approach harkens to the aspirational ``green button'' that would,
in response to a query, automatically retrieve patient data from an
institution's records with which to conduct on-demand retrospective
studies for an individual patient (Longhurst, Harrington, and Shah
2014). Several hurdles face the deployment of such an approach in
practice, but its benefits must first be demonstrated. Our motivations
in this scoping review are to describe the settings in and problems with
which this approach has been tasked, to attempt to organize them within
a common framework, and to evaluate the promise they show toward
achieving this or other needs in clinical informatics. Along the way, we
attempt to reconcile terminology, summarize motivations and evaluations,
and propose valuable follow-up work.

\hypertarget{related-work}{%
\subsection{Related work}\label{related-work}}

Clinical CBR emerged among rule-based approaches and other AI tools in
the development of expert systems (Aamodt and Plaza 1994). Early
implementations realized a general workflow described as ``the four
REs'', later the ``R4 cycle'' (Aamodt and Plaza 1994; Begum et al.
2011): Given a new case (or problem), the system \emph{retrieves} one or
more past cases (solved problems) from a corpus, \emph{reuses} these to
generate an understanding of (solution to) the new case, \emph{revises}
this understanding (solution) to better fit the new case (sometimes
called ``adaptation''), and \emph{retains} the new case and its eventual
understanding (solution) in the corpus to be retrieved and reused in
future. In contrast to rule-based systems, which are variable-based and
generally interpretable as a single rule applied to all new cases,
case-based systems provide case-specific interpretations in the form of
a number of more fully understood reference cases. While the similarity
measure used in the retrieval step need not be changed as the corpus
grows, proposed measures have been diverse, contested, and rarely
systematically compared.

Kolodner (1992) distinguished two styles of CBR: problem-solving, which
is more procedural and used when objectives are more clearly defined,
and interpretative, which provides categorizations and justifications
for possible solutions. A similar dichotomy is commonly used to
distinguish the performance criterion for the usefulness of predictive
models from the interpretability criterion for their usability. In these
terms, localized models are highly procedural: The step of fitting a
predictive model to a retrieved cohort is an automated adaptive strategy
(Begum et al. 2011), and the parameters that govern cohort retrieval can
be tuned alongside model hyperparameters in a machine learning (ML)
workflow. Accordingly, localized models have been primarily used for and
evaluated on their ability to predict classes or outcomes. Nevertheless,
as some recent applications have shown, they can be used to draw
inferences about the importance of different risk factors to specific
individuals. While most ML models come equipped with measures of feature
importance and model-agnostic tools can generate explanations for model
predictions, these quantifications are not directly interpretable model
components analogous to the split nodes of a decision tree (DT) or the
coefficients of a GLM. While the patient similarity measure used to
retrieve each cohort may be complicated, the cohort itself can be
directly inspected by the user. Provided the model family fitted to the
cohorts is interpretable, the localized model inherits this property.
Thus, localized models may embody both styles of CBR.

We refer the interested reader to several previous reviews of CBR in
medicine (Gierl, Bull, and Schmidt 1998; Begum et al. 2011; Choudhury
and Begum 2016), of measures of patient similarity (Dai, Zhu, and Liu
2020), and of uses of patient similarity in predictive models (Welch and
Kawamoto 2013; Sharafoddini, Dubin, and Lee 2017; Parimbelli et al.
2018). While the reviews of CBR focus on applications using health data,
the patient similarity reviews encompass many additional types of data
(various molecular -omics, genetic tests, medical images, laboratory
tests, patient preferences, patient-reported outcomes, tracking devices,
social media) and survey a much broader scope of models (exploratory
analysis via dimension reduction and cluster analysis; risk evaluation
and outcome prediction; clinical decision support and software tools).
For the present review, we are interested in how similarity matching on
patient-level health data is used to construct local cohorts for
predictive or inferential modeling.

\hypertarget{objectives}{%
\subsection{Objectives}\label{objectives}}

Our goals in this review are (1) to describe applications of
similarity-based localized models using patient-level health data and
(2) to provide a general framework for the design and evaluation of such
localized models. We focus narrowly on localized models, rather than
broadly on patient similarity--based clinical decision support tools, so
that we may thoroughly assess their value in terms of reported
evaluations and comparisons to other methods. This also allows us to
devise a high-level framework that specializes to the majority of these
methods, which we then use to frame our discussion and recommendations.

Our selection of relevant papers from the search corpus will be based on
the following inclusion/exclusion criteria:

\begin{itemize}
\tightlist
\item
  Uses case-level data from a corpus of past cases with known
  responses\newline (response may be outcome, diagnosis, subtype, etc.)
\item
  Defines a numeric multivariate case similarity measure\newline (allow
  integer-valued measures)
\item
  Uses the similarity measure to retrieve cohorts for index cases from
  the corpus\newline (for example, based on a training--testing
  partition)
\item
  Fits statistical models to cohorts from which to make predictions or
  draw inferences about index cases\newline (outcome predictions,
  survival estimates, risk factor contributions, model evaluation
  statistics, etc.)
\end{itemize}

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

Here we describe our review process, including deviations from plans and
the reasons for them. More details are included in Section
\ref{selection-process}.

\hypertarget{prisma-checklist}{%
\subsection{PRISMA checklist}\label{prisma-checklist}}

We include a PRISMA checklist as Supplemental Table 1 and a PRISMA
abstract checklist as Supplemental Table 2. Because we focus on
methodologies rather than conceptual approaches or evidence, we deviate
in some ways from PRISMA guidelines. In particular, those items of the
checklist involving bias assessment and quantitative synthesis are
intended for meta-analyses so did not apply to this study. Because we
are not aware of any standard procedures for conducting reviews and
syntheses of methodology, no protocol was prepared for this study.

\hypertarget{search}{%
\subsection{Search}\label{search}}

We derived the eligibility criteria itemized in the Introduction from a
seed set of previously read studies. Based on these criteria, we
formulated search strings for five platforms: PubMed, Web of Science,
Academic Search Premier, Google Scholar, and MathSciNet. We finalized
the PubMed search first, then adapted it to the other platforms (Section
\ref{search-design}).

Through each platform, we searched those databases included by default.
This means that we searched both MEDLINE and PubMed Central (PMC)
through Pubmed (we did not exclude other databases, but our earliest
included results post-date them) and that we searched the six indices of
the Core Collection (the Science Citation Index Expanded, the Social
Sciences Citation Index, the Arts \& Humanities Citation Index, the
Emerging Sources Citation Index, the Conference Proceedings Citation
Index, and the Book Citation Index) as well as several regional
databases through the Web of Science platform.

The structure and terms of our search strings were inspired in part by
previous reviews adjacent to our topic of interest (Sharafoddini, Dubin,
and Lee 2017; Parimbelli et al. 2018). We organized the PubMed search in
disjunctive normal form (an OR of ANDs). Following the solidification of
an outline, we expanded each term to include synonyms that are similar
enough to be applicable to our search. We then evaluated the expanded
search string using the PubMed Advanced Search platform. We initially
included each term and their synonyms separately to evaluate what
resulted. We pruned several terms that returned no results (``phrases
not found'') or to reduce the number of results. At the conclusions of
this process for each individual term, we combined the search terms
using disjunctive normal form.

@@

\hypertarget{titleabstract-screen}{%
\subsection{Title/abstract screen}\label{titleabstract-screen}}

Within Covidence, we screened titles and abstracts for relevance. We
decided on four eligibility criteria to expedite the screening process.
Articles must be written in English, for readability; they must be
original studies, to exclude secondary sources with duplicate
information; their use settings must be medical, clinical, or related,
to keep our review topical; and it must not be clear that their use of
our search terms was different from our intended meaning. Each of two
authors (AC and JCB) screened roughly half of the entries. They
regularly reviewed each other's decisions to improve consistency. In
cases of uncertainty, entries were included. For the update, one author
(PM) screened all results for three criteria and another (JCB) screened
the survivors for the fourth criterion.

\hypertarget{full-text-review}{%
\subsection{Full text review}\label{full-text-review}}

We set out 4 criteria for full-text review to restrict to studies that
used some form of localized modeling on health data:

\begin{itemize}
\tightlist
\item
  Pulls case-level data from a corpus of past cases with known classes
  or outcomes
\item
  Defines a numeric multivariate case similarity measure
\item
  Uses the similarity measure to select cohorts for index cases from the
  corpus
\item
  Fits statistical models to cohorts to make predictions or inferences
  about index cases
\end{itemize}

Note that classical CBR satisfies the first and third criteria by
definition and in most cases will satisfy the second.

In Covidence, two authors (AC and JCB) independently evaluated each
manuscript for these eligibility criteria. The first criterion that a
manuscript failed was designated the reason for exclusion. An antecedent
criterion was used to exclude manuscripts that did not report the
results of original studies involving real-world experiments or
empirical data, for example surveys of prior work and proposals of
frameworks. In cases of disagreement between the authors on the reason
for exclusion, the first criterion was adopted. The authors arrived at
agreement on inclusion or exclusion through discussion. We calculated
inter-rater reliability to evaluate our screening and review process.

During full text review, we decided to expand the conception of
statistical models (fourth criterion): Rather than restricting to models
that are fit and evaluated in separate steps, we chose to allow simple
statistical summaries such as mean survival (Mariuzzi et al. 1997) and
survival curves (Lowsky et al. 2013). These approaches were novel to CBR
and presage later developments, so were helpful to understanding the
development of localized modeling. However, this then admitted studies
that applied conventional nearest neighbors prediction: Each retrieved
cohort consisted of the \(k\) most similar cases to the index case, and
the response for the index case was predicted to be either the mean
(continuous response) or the plurality (discrete response) of the
cohort's responses. A review of all studies that use nearest neighbors
prediction would be impractical. Because our focus is on novel
approaches that combine similarity-based retrieval and statistical
modeling of retrieved cohorts, we chose to exclude those studies whose
approach was equivalent to nearest neighbors prediction using a
conventional similarity measure.

@@

\hypertarget{coding}{%
\subsection{Coding}\label{coding}}

@@

\hypertarget{synthesis}{%
\subsection{Synthesis}\label{synthesis}}

Rather than an evidence synthesis characteristic of most systematic
reviews, we here pursue a methodology synthesis to harmonize largely
independent research efforts that have converged on a common family of
techniques. The goal will be to describe a unified framework for
localized models that can be used to guide future study designs and
implementations as well as more systematically evaluate variations on
the theme and measure the dependence of results on modeling choices.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{selection}{%
\subsection{Selection}\label{selection}}

@@

\caption{PRISMA-S flow chart. Lowercase letters refer to items obtained from the seed set (m), the structured search (n), and citation/reference tracking (s). Within the structured search results, summands correspond to original and update.}\label{fig:prisma}
\end{figure}

>>>>

%Version 2.1 April 2023
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[sn-mathphys,Numbered,pdflatex]{sn-jnl}

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published
%%%%  by Springer Nature. The guidance has been prepared in partnership with
%%%%  production teams to conform to Springer Nature technical requirements.
%%%%  Editorial and presentation requirements differ among journal portfolios and
%%%%  research disciplines. You may find sections in this template are irrelevant
%%%%  to your work and are empowered to omit any such section if allowed by the
%%%%  journal you intend to submit to. The submission guidelines and policies
%%%%  of the journal take precedence. A detailed User Manual is available in the
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

%% Per the spinger doc, new theorem styles can be included using built in style, 
%% but it seems the don't work so commented below
%\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

%% \theoremstyle{thmstyletwo}%
\theoremstyle{remark}
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

%% \theoremstyle{thmstylethree}%
\theoremstyle{definition}
\newtheorem{definition}{Definition}%

\usepackage{lscape}
\newcommand{\landscapebegin}{\begin{landscape}}
\newcommand{\landscapeend}{\end{landscape}}

\usepackage{soul}

\raggedbottom

@@

\begin{document}

\title[Clinical prediction using similarity cohort--based
models]{Clinical prediction using similarity cohort--based models: A
systematic scoping review}

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

\author[1]{\fnm{Adi} \sur{Cohen} \dgr{BS}}

\author[2]{\fnm{Patti} \sur{McCall-Junkin} \dgr{MA, MLS}}

\author*[3]{\fnm{Jason
Cory} \sur{Brunson} \dgr{PhD}}\email{\href{mailto:jason.brunson@medicine.ufl.edu}{\nolinkurl{jason.brunson@medicine.ufl.edu}}}

  \affil[1]{\orgdiv{College of Osteopathic Medicine}, \orgname{Nova
Southeastern University}}
  \affil[2]{\orgdiv{Smathers Libraries Academic Research \& Consulting
Services}, \orgname{University of Florida}}
  \affil*[3]{\orgdiv{Laboratory for Systems
Medicine}, \orgname{University of
Florida}, \orgaddress{\city{Gainesville}, \country{USA}, \state{Florida}, \street{P.O.
Box 100225 JHMHC}}}

\abstract{\textbf{Background:} Applications of classical case-based
reasoning (CBR) have given rise to a family of modeling techniques
\hl{reliant on similarity-based cohorts}, in which a statistical model
is fitted to a neighborhood of labeled cases matched by similarity to a
target case.

\textbf{\hl{Objective:}} We aim to describe clinical and health
applications of \hl{similarity cohort--based models (SCBMs)} to date and
propose a general framework for their design and evaluation.

\textbf{Methods:} We \hl{conducted searches of} four bibliographic
platforms during 2021 July 19--22\hl{ and updated them on} 2024 January
24. We set four eligibility criteria to identify applications of
\hl{SCBMs} to clinical and health
tasks\hl{, which amounted to using a multivariate similarity measure to retrieve cases from a retrospective corpus relevant to new cases and fitting statistical models to these cohorts}.
Two authors divided title/abstract screening and reviewed screened
entries for inclusion. We discussed settings, tasks, and tools;
identified and tabulated themes; and synthesized the methods into a
general framework.

\textbf{Results:} Of 1,657 search results, 360 were reviewed, then
combined with 43 publications that seeded the review and 1 obtained by
citation tracking. 27 were included, published 1997--2022. The
specificity of search terms was poor\hl{ (51/1,657)}, and inter-rater
reliability was low\hl{ (82\%)}. Almost all models were predictive, the
most common tasks being prognosis\hl{ (11)} and diagnosis\hl{ (10)}.
Most studies used clinical\hl{ (22)}, occasionally laboratory\hl{ (7)}
or imaging\hl{ (2)}, data.
\hl{The most competitive performances were reported for models that further hybridized the basic approach.}
A general technique that specializes to most of those reviewed involved
matching, retrieval, fitting, and evaluation steps that could optionally
be supervised, optimized, or recursively performed.

\textbf{Conclusions:} \hl{SCBMs} have potential to improve the
performance of clinical decision support tools while maintaining
interpretability, but rigorous comparisons to competing methods must be
conducted and computational hurdles must be overcome. We hope that our
review will spur future work on efficiency, reproducibility, and user
needs.}

\keywords{case-based reasoning, similarity cohort, nearest
neighbors, clinical decision support, scoping review}

\pacs[JEL Classification]{C38, C65}
\pacs[MSC Classification]{62P10}

\maketitle

@@

\section{Background}\label{background}

A main driver of the development of clinical information systems (CIS)
and \hl{clinical decision support systems (CDSS)} has been to leverage
the advantages of routinely collected health data toward
\hl{improving care }\citep{Benchimol2015, Wasylewicz2019}. These
advantages include their immediate accessibility through the
institutional \hl{electronic health record (EHR)}, billing database, or
other source; their specificity to the institution and the population it
serves; and the closeness in time of the available data. \hl{Though not}
collected for research use, the secondary research use of routinely
collected health data has been put forth, as ``practice-based
evidence'', a complement to the paradigm of
\hl{evidence-based practice }\citep{Sim2001, Holmqvist2015}.

Over the same period of time, interest in individualizing care from
population-derived evidence-based recommendations to specific patient
needs has driven the application of
\hl{advanced computational tools }\citep{Rosella2022, Elhaddad2024},
including artificial intelligence (AI). Whereas classical models
produced formulae involving a limited set of data elements that would be
uniformly applied to all patients, AI models often process hundreds or
more variables in opaque ways to yield predictions that depend on not
only their values but
\hl{their associations and interplay with each other }\citep{Molnar2023}.
The loss of straightforward interpretations of these models has limited
their practical uptake and motivated the construction of explanatory
statistics for opaque models as well as the development of
\hl{more interpretable complex models }\citep{Rudin2022, Molnar2023}.

One of few methodologies to address all three of these established needs
is \hl{similarity cohort--based modeling (SCBM)}, which
\hl{we find in this review} to have been introduced several times
independently under different names, and in slightly different forms.
The approach is a specialized form of case-based reasoning (CBR) that
relies on a \hl{patient similarity measure }\citep{Brown2016} to extract
a cohort of past or training cases that then inform the diagnosis,
prognosis, or care of a new or test case.\footnote{Similarity measures
  between more granular units of analysis, such as encounters and
  decision points\hl{, or between patients' time series}, are often
  still referred to as patient similarity measures\hl{ in our sample}.}
\hl{Some of the earliest and simplest CBR tools returned} these
retrieved cases \hl{ to inform human decision-making}\citep{Aamodt1994};
the related nearest neighbors (NN) technique generates predictions from
cohorts automatically via averaging (regression) or voting
(classification) of their outcomes. In contrast, \hl{SCBM} fits families
of predictive models, for example generalized linear models (GLMs), to
retrieved similarity cohorts in order to generate predictions.
\hl{While many ML models come equipped with measures of feature importance and model-agnostic tools can generate explanations for model predictions }\citep{Molnar2023}\hl{, these quantifications are not directly interpretable model components analogous to the split nodes of a DT or the coefficients of a GLM.
While the patient similarity measure used to retrieve each cohort may be complicated, the cohort itself can be directly inspected.}

\hl{This approach offers an alternative paradigm for interpretable machine learning that combines the simplicity and transparency of classical models like decision trees (DTs) and GLMs with the sensitivity to nonlinearity and interactions of advanced opaque models like random forests (RFs) and artificial neural networks (ANNs). The sensitivity is enabled by the fitting of the model to the local cohort, whose size and scope are governed by hyperparameters that may be tuned like any others via grid or Bayesian optimization. Some similarity measures or other retrieval methods may be more interpretable than others, so that the interpretability of the SCBM approach would then depend on that of its retrieval (similarity matching) and revision (predictive modeling) steps. Moreover, this approach would benefit computationally as well as practically from the availability of a large corpus of cases, as efficient data structures and algorithms exist for similarity matching }\citep{Halder2024}\hl{ and the local models would benefit from increased numbers of relevant cases.}
\hl{SCBM}s thus provide an individualized way to fit fully interpretable
models to large population data.

This approach harkens to the aspirational ``green button'' that would,
in response to a query, automatically retrieve patient data from an
institution's records with which to conduct on-demand retrospective
studies for an individual patient \citep{Longhurst2014}. Several hurdles
face the deployment of such an approach in practice, but its benefits
must first be demonstrated. Our motivations in this scoping review are
to describe the settings in and problems with which this approach has
been tasked, to attempt to organize them within a common framework, and
to evaluate the promise they show toward achieving this or other needs
in clinical informatics. Along the way, we attempt to reconcile
terminology, summarize motivations and evaluations, and propose valuable
follow-up work.

\subsection{\texorpdfstring{Related
\hl{survey }work}{Related work}}\label{related-work}

Clinical CBR emerged among rule-based approaches and other AI tools in
the development of expert systems \citep{Aamodt1994}. Early
implementations realized a general workflow described as ``the four
REs'', later the ``R4 cycle'' \citep{Aamodt1994, Begum2011}: Given a new
case (or problem), the system \emph{retrieves} one or more past cases
(solved problems) from a corpus, \emph{reuses} these to generate an
understanding of (solution to) the new case, \emph{revises} this
understanding (solution) to better fit the new case (sometimes called
``adaptation''), and \emph{retains} the new case and its eventual
understanding (solution) in the corpus to be retrieved and reused in
future. In contrast to rule-based systems, which are variable-based and
generally interpretable as a single rule applied to all new cases,
case-based systems provide case-specific interpretations in the form of
a number of more fully understood reference cases.

\citet{Kolodner1992} distinguished two styles of CBR:
\emph{problem-solving}, which is more procedural and used when
objectives are more clearly defined, and \emph{interpretative}, which
provides categorizations and justifications for possible solutions. A
similar dichotomy is commonly used to distinguish the \emph{performance}
\hl{criterion }of predictive models from the \emph{interpretability}
criterion\hl{.} In these terms, \hl{SCBM}s are highly procedural: The
step of fitting a predictive model to a retrieved cohort is an automated
adaptive strategy \citep{Begum2011}, and the parameters that govern
cohort retrieval can be tuned alongside model hyperparameters in a
machine learning (ML) workflow. Accordingly, \hl{SCBM}s have been
primarily used for and evaluated on their ability to predict classes or
outcomes. Nevertheless, as \hl{we find in this review}, they can be used
to draw inferences about the importance of different risk factors to
specific individuals.

\hl{We also note that a large variety of AI tools, besides NN, use similarity measures or other relevance-based retrieval methods: Back-propagation in ANNs relies on differentiation, which requires that all network nodes' activation functions and, therefore, the final model are continuous and differentiable, i.e. that small distances between cases yield small changes in prediction; and kernel-based methods that rely directly on pairwise distances computed via inner product, rather than high-dimensional coordinates, extend far beyond their use in CBR. The key difference in SCBM is that the retrieved cohort is treated as a component of the model and made available to the user. Indeed, retrieval procedures were the most significant algorithmic requirements and accomplishments of CBR }\citep{Kolodner1992, Gierl1998}\hl{ and the retrieved cohorts may be reused for numerous modeling tasks.}
Thus, \hl{SCBM}s may embody both styles of CBR.

We refer the interested reader to several previous reviews of CBR in
medicine \citep{Gierl1998, Begum2011, Choudhury2016}, of measures of
patient similarity \citep{Dai2020}, and of uses of patient similarity in
predictive models \citep{Welch2013, Sharafoddini2017, Parimbelli2018}.
While the reviews of CBR focus on applications using health data, the
patient similarity reviews encompass many additional types of data
(various molecular -omics, genetic tests, medical images, laboratory
tests, patient preferences, patient-reported outcomes, tracking devices,
social media) and survey a much broader scope of models (exploratory
analysis via dimension reduction and cluster analysis; risk evaluation
and outcome prediction; \hl{CDSS} and software tools). For the present
review, we are interested in how similarity matching on patient-level
health data is used to construct local cohorts for predictive or
inferential
modeling.\hl{ While the similarity measure used in the retrieval step need not be changed as the corpus grows, proposed measures have been diverse, contested, and rarely systematically compared.}

\subsection{Objectives}\label{objectives}

Our goals in this review are (1) to describe applications of similarity
cohort--based models using patient-level health data and (2) to provide
a general framework for the design and evaluation of such \hl{SCBM}s. We
focus narrowly on \hl{SCBM}s, rather than broadly on patient
similarity--based clinical decision support tools, so that we may
thoroughly assess their value in terms of reported evaluations and
comparisons to other methods. This also allows us to devise a high-level
framework that specializes to the majority of these methods, which we
then use to frame our discussion and recommendations.

@@

\section{Methods}\label{methods}

Here we describe our review process, including deviations from plans and
the reasons for them. More details are included in \hl{the Appendix}.

\subsection{PRISMA checklist}\label{prisma-checklist}

We include a PRISMA checklist as Supplemental Table 1 and a PRISMA
abstract checklist as Supplemental Table 2. Because we focus on
methodologies rather than\hl{ evidence}, we deviate in some ways from
PRISMA guidelines. In particular, those items of the checklist involving
bias assessment and quantitative synthesis are intended for
meta-analyses so did not apply to this study. Because we are not aware
of any standard procedures for conducting reviews and syntheses of
methodology, no protocol was prepared for this study.

\subsection{Eligibility criteria}\label{eligibility-criteria}

\hl{Our selection of relevant papers from the search corpus will be based on the following inclusion/exclusion criteria:}

\begin{itemize}
\tightlist
\item
  \hl{Uses case-level data from a corpus of past cases with known responses}\newline
  \hl{(response may be outcome, diagnosis, subtype, etc.)}
\item
  \hl{Defines a numeric multivariate case similarity measure}\newline
  \hl{(allow integer-valued measures)}
\item
  \hl{Uses the similarity measure to retrieve cohorts for index cases from the corpus}\newline
  \hl{(for example, based on a training--testing partition)}
\item
  \hl{Fits statistical models to cohorts from which to make predictions or draw inferences about index cases}\newline
  \hl{(outcome predictions, survival estimates, risk factor contributions, model evaluation statistics, etc.)}
\end{itemize}

\hl{Note that classical CBR satisfies the first and third criteria by definition and in most cases will satisfy the second.}

\subsection{Search}\label{search}

We derived the eligibility \hl{criteria }from a seed set of previously
read studies. Based on these criteria, we formulated search strings for
five platforms: PubMed, Web of Science, Academic Search Premier, Google
Scholar, and MathSciNet. We finalized the PubMed search first, then
adapted it to the other platforms\hl{.}

@@

\subsection{Title/abstract screen}\label{titleabstract-screen}

\hl{We} screened titles and abstracts for relevance. We decided on four
eligibility criteria to expedite the screening process. Articles must be
written in English, for readability; they must be original studies, to
exclude secondary sources with duplicate information; their use settings
must be medical, clinical, or related, to keep our review topical; and
it must not be clear that their use of our search terms was different
from our intended meaning. Each of two authors (AC and JCB) screened
roughly half of the entries. They regularly reviewed each other's
decisions to improve consistency. In cases of uncertainty, entries were
included. For the update, one author (PM) screened all results for
\hl{the first }three criteria and another (JCB) screened the survivors
for the fourth criterion.

\subsection{Full text review}\label{full-text-review}

We \hl{used the} 4 criteria\hl{ from Section }\ref{eligibility-criteria}
for full-text review\hl{.}

\hl{Two} authors (AC and JCB) independently evaluated each manuscript
for these eligibility criteria. The first criterion that a manuscript
failed was designated the reason for exclusion. An antecedent criterion
was used to exclude manuscripts that did not report the results of
original studies involving real-world experiments or empirical data, for
example surveys of prior work and proposals of frameworks. In cases of
disagreement between the authors on the reason for exclusion, the first
criterion was adopted. The authors arrived at agreement on inclusion or
exclusion through discussion. We calculated inter-rater reliability to
evaluate our screening and review process.

@@

\subsection{Coding}\label{coding}

@@

\subsection{Synthesis}\label{synthesis}

Rather than an evidence synthesis characteristic of most systematic
reviews, we here pursue a methodology synthesis to harmonize largely
independent research efforts that have converged on a common family of
techniques. The goal will be to describe a unified framework for
\hl{SCBM}s that can be used to guide future study designs and
implementations as well as more systematically evaluate variations on
the theme and measure the dependence of results on modeling choices.

\section{Results}\label{results}

\subsection{Selection}\label{selection}

@@

\caption{\hl{PRISMA-S flow chart.}
 Lowercase letters refer to items obtained from
 the seed set (m),
 the structured search (n), and
 citation/reference tracking (s).
 Within the structured search results,
 summands correspond to original and update.}\label{fig:prisma}
\end{figure}
